---
title: "OpenFlights Analysis with grafeoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OpenFlights Analysis with grafeoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(grafeoR)
library(ggplot2)
```

The package ships with a compact real-world OpenFlights subset so you can test
the Rust binding against graph data that looks like an actual transport
network, not just toy nodes and edges.

The bundled sample contains:

- the 20 airports with the highest outbound route counts in the upstream
  OpenFlights snapshot
- the 1,129 airline route records whose source and destination are both inside
  that 20-airport subset

```{r load-data}
sample <- openflights_sample_data()

dim(sample$airports)
dim(sample$routes)
head(sample$airports[, c("iata", "city", "country", "snapshot_outbound_routes")])
head(sample$routes)
```

## Load the graph into Grafeo

```{r helpers}
gql_string <- function(x) {
  if (is.null(x) || is.na(x)) {
    return("NULL")
  }

  value <- gsub("\\\\", "\\\\\\\\", as.character(x), perl = TRUE)
  value <- gsub("\"", "\\\\\"", value, fixed = TRUE)
  paste0("\"", value, "\"")
}

load_openflights_graph <- function(db, sample) {
  tx <- db$begin()
  on.exit(
    if (tx$is_active()) {
      try(tx$rollback(), silent = TRUE)
    },
    add = TRUE
  )

  for (i in seq_len(nrow(sample$airports))) {
    row <- sample$airports[i, ]
    tx$execute(sprintf(
      paste0(
        "INSERT (:Airport {",
        "iata: %s, name: %s, city: %s, country: %s, ",
        "lat: %.6f, lng: %.6f, snapshot_outbound_routes: %d",
        "})"
      ),
      gql_string(row$iata),
      gql_string(row$name),
      gql_string(row$city),
      gql_string(row$country),
      row$latitude,
      row$longitude,
      as.integer(row$snapshot_outbound_routes)
    ))
  }

  for (i in seq_len(nrow(sample$routes))) {
    row <- sample$routes[i, ]
    tx$execute(sprintf(
      paste0(
        "MATCH (src:Airport {iata: %s}), (dst:Airport {iata: %s}) ",
        "INSERT (src)-[:ROUTE {airline: %s, stops: %d, equipment: %s}]->(dst)"
      ),
      gql_string(row$source_iata),
      gql_string(row$dest_iata),
      gql_string(row$airline),
      as.integer(row$stops),
      gql_string(row$equipment)
    ))
  }

  tx$commit()
  invisible(db)
}
```

```{r build-graph}
db <- grafeo_db()

load_openflights_graph(db, sample)
db$info()
```

## Query airport and route data back into R

For the charts below, `grafeoR` is used to pull nodes and edges back into R as
data frames, and the aggregation for plotting is handled on the R side.

```{r query-graph}
airports_tbl <- db$query(
  paste(
    "MATCH (a:Airport)",
    "RETURN a.iata, a.name, a.city, a.country, a.lat, a.lng,",
    "a.snapshot_outbound_routes",
    "ORDER BY a.snapshot_outbound_routes DESC, a.iata"
  )
)

routes_tbl <- db$query(
  paste(
    "MATCH (src:Airport)-[r:ROUTE]->(dst:Airport)",
    "RETURN src.iata, src.lat, src.lng, dst.iata, dst.lat, dst.lng, r.airline"
  )
)

route_segments <- aggregate(
  routes_tbl$r.airline,
  by = list(
    source_iata = routes_tbl$src.iata,
    source_lat = routes_tbl$src.lat,
    source_lng = routes_tbl$src.lng,
    dest_iata = routes_tbl$dst.iata,
    dest_lat = routes_tbl$dst.lat,
    dest_lng = routes_tbl$dst.lng
  ),
  FUN = length
)
names(route_segments)[names(route_segments) == "x"] <- "airline_count"
route_segments <- route_segments[
  order(
    -route_segments$airline_count,
    route_segments$source_iata,
    route_segments$dest_iata
  ),
  ,
]

head(airports_tbl[, c("a.iata", "a.city", "a.country", "a.snapshot_outbound_routes")])
head(route_segments)
```

## Visualize the hubs

```{r top-hubs-plot, fig.width=8, fig.height=5}
top_hubs <- airports_tbl[seq_len(min(10L, nrow(airports_tbl))), , drop = FALSE]
top_hubs$airport <- factor(
  paste(top_hubs$a.iata, top_hubs$a.city, sep = " - "),
  levels = rev(paste(top_hubs$a.iata, top_hubs$a.city, sep = " - "))
)

ggplot(
  top_hubs,
  aes(
    x = airport,
    y = a.snapshot_outbound_routes,
    fill = a.country
  )
) +
  geom_col(width = 0.75, color = "white") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Top OpenFlights Hubs in the Bundled Snapshot",
    subtitle = "The subset is small enough for a fast local demo, but still looks real",
    x = NULL,
    y = "Snapshot outbound routes",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Visualize the route network

```{r route-map-plot, fig.width=10, fig.height=6}
labels <- airports_tbl[seq_len(min(10L, nrow(airports_tbl))), , drop = FALSE]

ggplot() +
  geom_segment(
    data = route_segments,
    aes(
      x = source_lng,
      y = source_lat,
      xend = dest_lng,
      yend = dest_lat,
      linewidth = airline_count,
      alpha = airline_count
    ),
    color = "#0f4c5c",
    lineend = "round"
  ) +
  geom_point(
    data = airports_tbl,
    aes(
      x = a.lng,
      y = a.lat,
      size = a.snapshot_outbound_routes
    ),
    shape = 21,
    fill = "#ffbf69",
    color = "#7f5539",
    stroke = 0.4
  ) +
  geom_text(
    data = labels,
    aes(
      x = a.lng,
      y = a.lat,
      label = a.iata
    ),
    size = 3,
    nudge_y = 3,
    check_overlap = TRUE
  ) +
  coord_quickmap() +
  scale_linewidth(range = c(0.2, 1.3), guide = "none") +
  scale_alpha(range = c(0.15, 0.7), guide = "none") +
  scale_size(range = c(2.5, 7), guide = "none") +
  labs(
    title = "Airline Routes Within the Bundled OpenFlights Top-20 Subgraph",
    subtitle = "Line width reflects the number of airline records between each airport pair",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f8f4ec", color = NA),
    panel.background = element_rect(fill = "#f8f4ec", color = NA)
  )
```

The bundled sample is documented in
`inst/extdata/openflights-README.md`, including the upstream source URLs and
license attribution to OpenFlights.

```{r cleanup, include = FALSE}
db$close()
```
